---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# {rgeowheels}

<!-- badges: start -->
<!-- badges: end -->

 Downloads pre-compiled Windows 'wheel' files (.whl) for Python geospatial 
  packages from <https://github.com/cgohlke/geospatial-wheels>. These are unofficial 
  binary installers for some Python geospatial libraries on Windows prepared by 
  Christoph Gohlke. Wheels for various packages, Python versions, and architectures
  are made available via GitHub releases, providing the ability to revert to
  prior versions when needed.

## Installation

You can install the development version of {rgeowheels} like so:

``` r
# install.packages("remotes")
remotes::install_github("brownag/rgeowheels")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(rgeowheels)

if (Sys.info()["sysname"] == "Windows")
  install_wheel("GDAL")

install_wheel("rasterio", pyversion = "3.8", url_only = TRUE)
```
